A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-08-10 docs/abstract_v1.01.tex
Made notes for the introduction.
Wrote more stuff.

We analyse the stellar and dark matter distribution in the spiral galaxy SDSS J1331+3638 (J1331) by means of two independent methods: gravitational lensing and dynamical Jeans modelling. Hubble Space Telescope (HST) imaging by \citet{SWELLSI} reveals, that J1331's bulge is superimposed by a quadruplet of extended lensing images. By fitting a gravitational potential model to the image positions, we constrain the mass inside the Einstein radius ($R_\text{ein}=0.91\pm0.02$ arcsec) to within 4\% ($M_\text{ein} = (7.8\pm0.3) \cdot 10^{10} M_\odot$). From Multi-Gaussian Expansions (MGE) of J1331's surface brightness distribution we find that J1331 has an apparent I-band magnitude of $m_I \simeq 15.8$ mag, a total luminosity of $L_{I,tot} \simeq 5.6 \cdot 10^{10}L_{I,\odot}$ and an effective radius of $R_\text{eff} \simeq 2.6$ arcsec = 5.6 kpc.

@@

>>>>

We analyse the stellar and dark matter distribution in the spiral galaxy SDSS J1331+3638 (J1331) by means of two independent methods: gravitational lensing and dynamical Jeans modelling. Hubble Space Telescope (HST) imaging by \citet{SWELLSI} reveals, that J1331's bulge is superimposed by a quadruplet of extended lensing images. By fitting a gravitational potential model to the image positions, we constrain the mass inside the Einstein radius ($R_\text{ein}=0.91\pm0.02$ arcsec) to within 4\% ($M_\text{ein} = (7.8\pm0.3) \cdot 10^{10} M_\odot$). From Multi-Gaussian Expansions (MGE) of J1331's surface brightness distribution we find that J1331 has a total I-band luminosity of $L_{I,tot} \simeq 5.6 \cdot 10^{10}L_{I,\odot}$ and an effective radius of $R_\text{eff} \simeq 2.6$ arcsec = 5.6 kpc. [TO DO: I guess, apparent magnitude is not very interesting.]

@@

[TO DO: something about mass to light ratios????]

### 2015-08-10 docs/results_lensing_v1.01.tex
Wrote more stuff.

@@

\paragraph{Comparison with Light Distribution.} Fig. \ref{fig:lenscomparemass} shows the surface mass distribution as predicted by the best fit model in tab. \ref{tab:bestfitlensmodel}. We introduce random noise according to the uncertainties in the Fourier shape parameters to create a mock observation that visualizes the effect of the measurement errors. To be able to compare the predicted mass distribution to the observed light distribution we transform the surface brightness into a mass density: We first integrate the MGE in tab. \ref{tab:MGEF814W} to get the total luminosity within the Einstein radius and the predicted mass-to-light ratio as $\Upsilon_\text{I,ein} = M_\text{ein} / L_\text{I,ein}$. Fig. \ref{fig:lenscomparelight} shows then the observed surface brightness in the F814W filter multiplied by $\Upsilon_\text{I,ein}$.  Fig. \ref{fig:lenscompareboth} finally compares equidensity contours at the same values of both the predicted lens mass distribution and the observed surface brightness times $\Upsilon_\text{I,ein}$. We note the following three facts: 1.The mass predicted from lensing and the observed light distribution are oriented in the same direction. 2. Within the Einstein radius mass and light distribution have the same shape, while further out the mass distribution is more roundish. 3. The light distribution drops faster than the mass with increasing radius. This could indicate that the stellar component of the galaxy is superimposed by a more roundish dark matter component.

@@

\begin{figure*}
\centering
\begin{subfigure}{.3\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_surface_brightness.ps}
  \caption{???}
  \label{fig:lenscomparelight}
\end{subfigure}%
\begin{subfigure}{.3\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_surface_density.ps}
  \caption{???}
  \label{fig:lenscomparemass}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/paper_lensmass_brightness_compare.ps}
  \caption{???}
  \label{fig:lenscompareboth}
\end{subfigure}
\caption{??? Preliminary crappy caption: Left: OBSERVED surface bightness multiplied with M/L in Einstein radius (overplotted in red). Middle: BEST FIT MODEL for mass distribution from lensing (including "wriggles" due to uncertainties in image positions). Contours are at the same levels. Right: Same contours, to directly show the difference in shape. ??? [TO DO: nice caption]}
\label{fig:???}
\end{figure*}

>>>>

\paragraph{Image Positions.} We determine the positions of the lensing images by first subtracting a smooth model for the galaxy's surface brightness from the original image. As models we use MGE fits (cf. \S\ref{kap:MGE}) and IRAF ellipse fits (???) to the galaxy in each the F450W and F814W filter. (For example the MGE we use for F814W is the MGE given in tab. \ref{tab:MGEF814W} convolved with the PSF in tab. \ref{tab:PSFMGEF814W}.) The lensing images become then visible in the residuals (see fig. \ref{fig:lens_just_imgpos}). Because the lensing images are extended, we use the position of the brightest pixel in each of the images. We also use the F814W-MGE subtracted residuals from \citet{SWELLSIII}. The lensing positions as determined from the latter are given in tab. \ref{tab:lenspos}. The scatter of lensing positions as determined from subtracting different surface brightness models from the galaxy in different filters gives an error of $\pm 1$ pixel on the image positions.

\begin{table}
\centering
\begin{minipage}{70mm}
\begin{tabular}{r|rrrr|c}
\hline
  & A & B & C & D & G\\\hline
$x_i$ [pixel] & 12.1 & -8.5 & 21.7 & -3.3 & 0.5 $\pm \sqrt{2}$ \\
$y_i$ [pixel] & 16.6 & -10.4 & -0.5 & 19.2 & 0.5 $\pm \sqrt{2}$ \\
\hline
\end{tabular}
\caption{Positions of the lensing images (A-D) and the galaxy center (G) in fig. \ref{fig:lens_just_imgpos}. The image positions were determined from the lens subtracted image for J1331 in figure 4 of \citet{SWELLSIII}, rotated to the $(x,y)$ coordinate system in fig. \ref{fig:lens_just_imgpos}. The pixel scale is 1 pixel = 0.05 arcsec and the error of each image position is $\pm$ 1 pixel. \textit{SMALL PROBLEM: Somehow I used $\sqrt(2)$ pixel as the error on the galaxy center in the Monte Carlo sampling code instead of the 0.5 pixel I claim here. Should I change this table and the error bars in the figures to match this bug????}}
\label{tab:lenspos}
\end{minipage}
\end{table}

@@

\paragraph{Comparison with Light Distribution.} Fig. \ref{fig:lenscomparemass} shows the surface mass distribution as predicted by the best fit model in tab. \ref{tab:bestfitlensmodel}. We introduce random noise according to the uncertainties in the Fourier shape parameters to create a mock observation that visualizes the effect of the measurement errors. From the mock image's second moment we find an average axis ratio for the lens mass model of $q_\text{lens} \simeq 0.695$, which is consistent with the one found by \citet{SWELLSIII}, $q_\text{lens,SWELLSIII} = 0.67 \pm 0.09$, while the light's average axis ratio in Table \ref{tab:galaxyparameters} is $q' = 0.598$.
\\To be able to compare the predicted mass distribution to the observed light distribution, we transform the surface brightness into a mass density: We first integrate the MGE in tab. \ref{tab:MGEF814W} to get the total luminosity within the Einstein radius and the predicted mass-to-light ratio as $\Upsilon_\text{I,ein} = M_\text{ein} / L_\text{I,ein}$. Fig. \ref{fig:lenscomparelight} shows then the observed surface brightness in the F814W filter multiplied by $\Upsilon_\text{I,ein}$.  Fig. \ref{fig:lenscompareboth} finally compares equidensity contours at the same values of both the predicted lens mass distribution and the observed surface brightness times $\Upsilon_\text{I,ein}$.
\\Figure \ref{fig:lenslightcompareALL} leads to the following three findings: 1.The mass predicted from lensing and the observed light distribution are oriented in the same direction. 2. Within the Einstein radius, mass and light distribution have the same shape, while further out the mass distribution is more roundish. 3. The light distribution drops faster than the mass with increasing radius. 
\\Reasons for the difference in observed light and measured mass distribution could be, e.g. an apparent change of shape due to dust extinction, a strongly changing $\Upsilon_*$, or the stellar component of the galaxy could be superimposed by a more roundish dark matter component.

@@

\begin{figure*}
\centering
\begin{subfigure}{.3\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_surface_brightness.ps}
  \caption{???}
  \label{fig:lenscomparelight}
\end{subfigure}%
\begin{subfigure}{.3\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/lens_surface_density.ps}
  \caption{???}
  \label{fig:lenscomparemass}
\end{subfigure}
\begin{subfigure}{.3\textwidth}
  \centering
  \includegraphics[width=.9\linewidth]{fig/paper_lensmass_brightness_compare.ps}
  \caption{???}
  \label{fig:lenscompareboth}
\end{subfigure}
\caption{??? Preliminary crappy caption: Left: OBSERVED surface bightness multiplied with M/L in Einstein radius (overplotted in red). Middle: BEST FIT MODEL for mass distribution from lensing (including "wriggles" due to uncertainties in image positions). Contours are at the same levels. Right: Same contours, to directly show the difference in shape. ??? [TO DO: nice caption]}
\label{fig:lenslightcompareALL}
\end{figure*}

### 2015-08-10 docs/results_photometry_v1.01.tex
Wrote more stuff.

@@

\paragraph{Transformation into physical units.} To transform the MGE in units of counts into physical units, we apply a simplified version of the procedure described in \citet{Holtzman}.
\\The scaling of the drizzled HST/WFC3 images is  $S = 0.05$ arcsec/pixel and the total exposure time $T = 1600$ sec. The total F814W luminosity in counts of each Gaussian of the MGE has a central flux in counts per pixel of
\begin{equation*}
C_0\text{[counts/pixel]} = \frac{L[counts]}{2\pi \sigma[\text{pixel}]^2 q}.
\end{equation*}
This is then transformed into an I-band surface brightness via
\begin{equation*}
\mu_I \simeq -2.5 \log_{10}\left( \frac{C_0\text{[counts/pixel]}}{T \cdot S^2}\right) + Z + C + A_I,
\end{equation*}
where $Z\simeq21.62$ is a the zero-point from \citet{Holtzman}  (updated according to \citet{Dolphin,DolphinNew}) for the photometric system of the HST/WFPC2 camera and the F814W filter plus a correction for the difference in gain of calibration and observation. $C= 0.1$ corrects for the finite aperture of the WFPC2. And $A_I =0.015$ mag  is the extinction in the I-band tpwards J1331, according to the NASA/IPAC Extragalactic Database [TO DO: proper reference]. The correction between the F814W filter and the I-band of the UBVRI photometric system is  small \citep{Holtzman} and we neglect it therefore.

[TO DO]

\paragraph{[TO DO: Stuff to mention]}
\begin{itemize}
\item fit sof MGE to J1331s surface brightness within $\sim$ 2 half-light radii, and to an Iraf Ellipse model for J1331's outer regions, make it possible to easily deconvolve the hst images with its PSF.
\item the deporjecten if the galaxy under the assumption of oblate axisymmetry and an estimated inclination of $\sim70^\circ$ can be performed analytically.
\end{itemize}

\begin{figure}
\begin{minipage}[c]{\linewidth}
\centering
\includegraphics[width=0.8\textwidth]{fig/1331F814Wsci_MGE_M.ps}
\caption{MGE for J1331's inner regions: Comparison of contours with constant F814W surface brightness of J1331's central region (orange lines) with the corresponding iso-brightness contours of the best fit MGE in table \ref{tab:MGEF814W}, convolved with the PSF in table \ref{tab:MGEF814W}, (gray lines). The MGE model is a good representation of the galaxy's light distribution within $\sim 5$ arcsec. Image scaling and orientation are indicated in the figure. The black line has a length of 10 arcsec and it's orientation corresponds to the galaxy's position angle as found in table \ref{tab:galaxyparameters}. For comparison the Einstein radius as found in table ??? is indicated as red dashed line. This MGE is used in the dynamical Jeans modelling in \S ???. [TO DO: explain how this MGE is used in dynamical modelling]}
\label{fig:MGEinnerRegions}
\end{minipage}
\begin{minipage}[c]{\linewidth}
\centering
\includegraphics[width=0.8\textwidth]{fig/1331F814W_MGE_disk_L.ps}
\caption{MGE for J1331's outer regions: Comparison of contours with constant surface brightness of the smooth IRAF ELLIPSE model for J1331 in the F814W filter (yellow) with a corresponding best fit MGE (gray lines). The green box corresponds to the image section shown in Fig. \ref{{fig:MGEinnerRegions}, with the Einstein radius as dashed red line. [TO DO: caption]}
\label{fig:MGEouterRegions}
\end{minipage}
\end{figure}

\begin{table*}
\centering
\begin{minipage}[t]{40mm}
\caption{F814W PSF MGE: Parameters of the circular four-Gaussian MGE in Eq. (\ref{eq:PSFgeneral} fitted to the radial profile of the synthetic HST/F814W PSF image by [TO DO].}
\begin{tabular}{ccc}
\hline
$k$ & $G_k$ & $\delta_k$ [arcsec] \\\hline
1 & 0.184 & 0.038\\
2 & 0.485 & 0.085\\
3 & 0.222 & 0.169\\
4 & 0.109 & 0.487\\\hline
\end{tabular}
\label{tab:PSFMGEF814W}
\end{minipage}
\hspace{10mm}
\begin{minipage}[t]{100mm}
\caption{J1331's F814W MGE: Parameters of the best fit MGE to the F814W surface brightness of J1331 in Fig. \ref{fig:F814W}. The fit is best inside an radius of 5 arcsec. The galaxy center and position angle, which gives the orientation of the MGE with respect to the original image, are given in table \ref{tab:galaxyparameters}. This MGE is used in the dynamical modelling in \S ???.}
\begin{tabular}{cccccc}
\hline
 & total luminosity  & surface density & \multicolumn{2}{c}{Gaussian dispersion} & axis ratio\\
$k$  & $L_k$ [counts] & $I_{0,k}$ [$L_\odot$/pc$^2$] & $\sigma_k$ [arcsec] & $\sigma_k$ [kpc] & $q'_k$\\\hline
1  &     9425.96 &      20768.  &  0.051   & 0.103  & 1.00\\
2  &    13173.0 &        3161.2 &  0.178   & 0.358  & 0.76\\
3  &    40235.0 &        1588.2 &  0.503   & 1.008  & 0.58\\
4  &    67755.2 &         502.25&  1.180   & 2.368  & 0.56\\
5  &    203677. &         136.51&  3.891   & 7.805  & 0.57\\\hline
\end{tabular}
\label{tab:MGEF814W}
\end{minipage}
\end{table*}

\begin{table}
\centering
\begin{minipage}{140mm}
\caption{Galaxy Parameters of J1331}
\begin{tabular}{lllrl}
\hline
redshift                  & $z_d$ & 0.113 & \citep{SWELLSIII}\\
angular diameter distance & $D_d$ [Mpc] & 414 & \\
scaling                   & 1 kpc / 1 arcsec & 2.006 & \\
position angle            & $\phi$ [degrees] & wrt what???\\
average axis ratio & $q'$ & 0.598\\
average ellipticity & $\epsilon = 1 - q'$ & 0.402 & \\
apparent F814W magnitude & $m_\text{F814W}$ [mag] & 15.77 & \\
total F814W luminosity & $L_\text{tot,F814W}$ [$10^{10} L_\odot$] & 5.6 & \\
effective half-light radius & $R_\text{eff}$ [arcsec] & 2.6 & \\
& $R_\text{eff}$ [pc]& 5.2 & \\
\hline
\end{tabular}
\label{tab:galaxyparameters}
\end{minipage}
\end{table}

>>>>

\begin{figure}
\centering
\includegraphics[width=0.8\columnwidth]{fig/1331F814Wsci_MGE_M.ps}
\caption{MGE for J1331's inner regions: Comparison of contours with constant F814W surface brightness of J1331's central region (orange lines) with the corresponding iso-brightness contours of the best fit MGE in table \ref{tab:MGEF814W}, convolved with the PSF in table \ref{tab:MGEF814W}, (gray lines). The MGE model is a good representation of the galaxy's light distribution within $\sim 5$ arcsec. Image scaling and orientation are indicated in the figure. The black line has a length of 10 arcsec and it's orientation corresponds to the galaxy's position angle as found in table \ref{tab:galaxyparameters}. For comparison the Einstein radius as found in table ??? is indicated as red dashed line. This MGE is used in the dynamical Jeans modelling in \S ???. [TO DO: explain how this MGE is used in dynamical modelling]}
\label{fig:MGEinnerRegions}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.8\columnwidth]{fig/1331F814W_MGE_disk_L.ps}
\caption{MGE for J1331's outer regions: Comparison of contours with constant surface brightness of the smooth IRAF ELLIPSE model for J1331 in the F814W filter (yellow) with a corresponding best fit MGE (gray lines). The green box corresponds to the image section shown in Fig. \ref{fig:MGEinnerRegions}, with the Einstein radius as dashed red line. [TO DO: caption]}
\label{fig:MGEouterRegions}
\end{figure}

\begin{table}
\centering
\caption{F814W PSF MGE: Parameters of the circular four-Gaussian MGE in Eq. (\ref{eq:PSFgeneral} fitted to the radial profile of the synthetic HST/F814W PSF image by [TO DO].}
\begin{tabular}{ccc}
\hline
$k$ & $G_k$ & $\delta_k$ [arcsec] \\\hline
1 & 0.184 & 0.038\\
2 & 0.485 & 0.085\\
3 & 0.222 & 0.169\\
4 & 0.109 & 0.487\\\hline
\end{tabular}
\label{tab:PSFMGEF814W}
\end{table}

\begin{table*}
\centering
\caption{J1331's F814W MGE: Parameters of the best fit MGE to the F814W surface brightness of J1331 in Fig. \ref{fig:F814W}. The fit is best inside an radius of 5 arcsec. The galaxy center and position angle, which gives the orientation of the MGE with respect to the original image, are given in table \ref{tab:galaxyparameters}. This MGE is used in the dynamical modelling in \S ???.  The first column gives the total F814W luminosity of the Gaussian in Eq. (\ref{eq:centralItotalL}) in units of counts. The second column is the corresponding I-band peak surface brightness in Eq. (\ref{eq:MGEgeneral}) in units of a luminosity surface density, calculated from the first column following the procedure described in \S ???. The third and fourth column give the dispersion and the last column the axis ratio of the Gaussian in Eq. (\ref{eq:MGEgeneral}).}
\begin{tabular}{cccccc}
\hline
 & total luminosity  & surface density & \multicolumn{2}{c}{Gaussian dispersion} & axis ratio\\
$k$  & $L_k$ [counts] & $I_{0,k}$ [$L_\odot$/pc$^2$] & $\sigma_k$ [arcsec] & $\sigma_k$ [kpc] & $q'_k$\\\hline
1  &     9425.96 &      20768.  &  0.051   & 0.103  & 1.00\\
2  &    13173.0 &        3161.2 &  0.178   & 0.358  & 0.76\\
3  &    40235.0 &        1588.2 &  0.503   & 1.008  & 0.58\\
4  &    67755.2 &         502.25&  1.180   & 2.368  & 0.56\\
5  &    203677. &         136.51&  3.891   & 7.805  & 0.57\\\hline
\end{tabular}
\label{tab:MGEF814W}
\end{table*}

\begin{table*}
\centering
\caption{Galaxy Parameters of J1331}
\begin{tabular}{lllrl}
\hline
redshift                  & $z_d$ & 0.113 & \citep{SWELLSIII}\\
angular diameter distance & $D_d$ [Mpc] & 414 & \\
scaling                   & 1 kpc / 1 arcsec & 2.006 & \\
position angle            & $\phi$ [degrees] & wrt what???\\
average axis ratio & $q'$ & 0.598\\
average ellipticity & $\epsilon = 1 - q'$ & 0.402 & \\
apparent I-band magnitude & $m_\text{I}$ [mag] & 15.77 & \\
total I-band luminosity & $L_\text{tot,I}$ [$10^{10} L_\odot$] & 5.6 & \\
effective half-light radius & $R_\text{eff}$ [arcsec] & 2.6 & \\
& $R_\text{eff}$ [pc]& 5.2 & \\
\hline
\end{tabular}
\label{tab:galaxyparameters}
\end{table*}

@@

\paragraph{Transformation into physical units.} To transform the MGE in units of counts into physical units, we apply a simplified version of the procedure described in \citet{Holtzman}, analogous to [TO DO: Cappellari Read me file???]
\\The scaling of the drizzled HST/WFC3 images is  $S = 0.05$ arcsec/pixel and the total exposure time $T = 1600$ sec. The total F814W luminosity in counts of each Gaussian of the MGE has a central surface brightness in counts per pixel of
\begin{equation*}
C_0\text{[counts/pixel]} = \frac{L[\text{counts}]}{2\pi \sigma[\text{pixel}]^2 q}.
\end{equation*}
This is then transformed into an I-band surface brightness via
\begin{equation}
\mu_I \simeq -2.5 \log_{10}\left( \frac{C_0\text{[counts/pixel]}}{T[\text{sec}] \cdot S[\text{arcsec/pixel}]^2}\right) + Z + C + A_I, \label{eq:muI_}
\end{equation}
where $Z\simeq21.62$ is a the zero-point from \citet{Holtzman}  (updated according to \citet{Dolphin,DolphinNew}) for the photometric system of the HST/WFPC2 camera and the F814W filter plus a correction for the difference in gain between calibration and observation. $C= 0.1$ corrects for the finite aperture of the WFPC2. And $A_I =0.015$ mag  is the extinction in the I-band tpwards J1331, according to the NASA/IPAC Extragalactic Database [TO DO: proper reference]. The color-dependent correction between the F814W filter and the I-band of the UBVRI photometric system is  small \citep{Holtzman} and we neglect it therefore. The last step is to transform the surface brightness $\mu_i$ in mag to the I-band surface density of the Gaussian in $L_\odot$/pc$^2$ as
\begin{eqnarray*}
I_0[L_\odot \text{pc}^{-2}] = \left( 64800/\pi\right)^2 \left(1+z \right)^4 10^{0.4\left(M_{\odot,I}-\mu_I \right)},
\end{eqnarray*}
where the term with $z$ accounts for redshift dimming and $M_{\odot,I}=4.08$ mag is the sun's absolute I-band magnitude \citep{1998gaas.book.....B}. The luminosity $L_k$[counts] and the corresponding surface brightness density $I \equiv I_{0,k}[L_\odot \text{pc}^{-2}]$ of each Gaussian are given in Table \ref{fig:MGEinnerRegions}. [TO DO: I'm really confused about this. Check, that everything is correct and the naming of quantities, e.g. with or without subscripts of 0, is fine???] [TO DO: Maybe shift to appendix???]

\paragraph{Inclination.} [TO DO: Do we need this for the dynamical modelling???? Is described on p. 56]

\paragraph{Total luminosity and effective radius.} J1331's total I-band luminosity is easily determined by summing up the luminosity contributions of all the Gaussians of the MGE for the outer regions (shown as gray lines in Fig. \ref{fig:MGEouterRegions}). We find $L_\text{tot,I} \simeq 5.6 \cdot 10^{10} L_\odot$. This corresponds to an apparent magnitude of $m_I = 15.77$ mag. We determine the circularized effective radius $R_\text{eff}$ of J1331 from the definition $L(<R_\text{eff}) \equiv \frac 12 L_\text{tot}$ and the growth curve $L(<R)$ from the MGE model of the outer regions [TO DO: Maybe I need a table anyway???], where $R$ is the projected radius on the sky [TO DO: Or is it $R'$???]. We find the effective radius to be $R_\text{eff} \simeq 2.6 \text{arcsec} \hat{=} 5.2 \text{kpc}$.  All values are summarized in Table \ref{tab:galaxyparameters}.

\paragraph{[TO DO: Stuff to mention]}
\begin{itemize}
\item the deporjecten if the galaxy under the assumption of oblate axisymmetry and an estimated inclination of $\sim70^\circ$ can be performed analytically.
\end{itemize}
